---
title: Introduction to quantitative time diary analysis
subtitle: Day 2 exercices
author: Pierre Walthéry
organisation:   UK Data Service and Centre for Time Use Research 
date: "17 October 2023"
mainfont: Arial
sansfont: Arial
#fig-width: 90
#fig-height: 60


format:
  pdf:
    toc: false
#    number-sections: false
    colorlinks: true
    papersize: A4
    geometry:
        - top=20mm
        - left=20mm
        - right=20mm
        - bottom=20mm
#        - heightrounded
#  html:
#    toc: true
pdf-engine: lualatex
execute:
  echo: true
  warning: false
---


```{r prep}
rm(list=ls())
library(dplyr)
library(ggplot2)
library(haven)

library(reshape2)
setwd("C://Users/qtnvpw1/Dropbox/work/UKDS/trainings/TU2023/Day_1")
```

##   1. Another illustration of computing episodes from episodes files
We will be using MTUS data from last week


Let's start by opening the episode and aggregate (day level)  files as we did last week

```{r open}
ep<-read_dta("data/mtus_teach_ep.dta")%>%
    select(-msamp,-swave,-ict,-core25,-diary,-inout)%>%
  filter(child!=1)

d<-read_dta("data/mtus_teach_ind.dta")
```
As before we create the STUDY variable in order to make our life easier 
```{r study}
d$study<-paste(d$country,d$survey,sep=" ")
ep$study<-paste(ep$country,ep$survey,sep=" ")
```

This time, we will be looking at modes of travel.
In the MTUS travelling is coded by reason for travel

Using the same method as last time, we create a variable flagging travel for any reason (MTUS codes 62 to 68) undertaken either by car (MTRAV = 1) or any other mode.

Note the same approach may be used to exploit information about location or copresence in time estimates.

In the code below TRC.T records the time spent travelling by car...
And TRC.O, by any other mean...

```{r trc}
ep<-ep%>%mutate(trc.t=ifelse(main>=62 & main<=68 & mtrav==1,time,0),
                tro.t=ifelse(main>=62 & main<=68 & mtrav>1  & mtrav<=5,time,0))

```

We can then compute the aggregate amounts of time of travel by car and other means for each day

```{r daytrav}
ep<-ep%>%group_by(study,hldid,persid,id)%>%mutate(trc.d=sum(trc.t),
                                                  tro.d=sum(tro.t))%>%ungroup()
```
Let's have a quick look at the estimates

```{r desc}
hist(ep$trc.d[ep$epnum==1 & ep$trc.d>0],freq=F)
hist(ep$tro.d[ep$epnum==1 & ep$tro.d>0],freq=F)

ep%>%filter(epnum==1)%>%summarise(car=mean(trc.d),
                                  oth=mean(tro.d))
```


We can also  examine the mean and median duration  of  car journeys  vs journeys with other means of transport by country. 

Spain seems to be an outlier, with fewer car journeys (duration=0) than other countries. The US have the longest daily duration of car journeys, whereas the Netherlands have the longest duration of non car journeys, perhaps due the importance of travel by bicycle in the country.
```{r country}
ep%>%filter(epnum==1)%>%
     group_by(study)%>%
     summarise(car.mn=mean(trc.d),
               car.md=median(trc.d),
               oth.mn=mean(tro.d),
               oth.md=median(tro.d))
```

We can also  examine the probability of travelling by car vs other means of transport by country. Please note that these are probabilities that a journey undertaken on a given day involves a car, or other means of transport. NOT the probability for someone to travel by car/other means.

```{r probtrav}
ep%>%filter(epnum==1)%>%
     group_by(study)%>%
     summarise(prc=mean(trc.d>0),pro=mean(tro.d>0))
```
... As well as looking at differences according by gender.  As done previously, we need to add the daily totals from the episode dataset to the dataset that contains person-level information. Here, the aggregate dataset

```{r travgend}
dt<-merge(d,
          ep%>%filter(epnum==1)%>%select(study,hldid,persid,id,trc.d,tro.d),
          by=c("study", "hldid","persid","id"),
          all.x=F,all.y=F)
```

Let's create a more explicit gender variable

```{r gendvar}
dt$gender<-as_factor(dt$sex)
levels(dt$gender)<-c("Male","Female")
```

And finally produce the median estimates of travel duration by gender. We can  also compute the ratio between the two. 

There are interesting gender differences, with men spending more time travelling by car than women across the board in Spain, France and the Netherlands. In most cases people spend more time overall travelling by other means than car, except, for the US.

```{r gendtrav}
dt%>%group_by(study,gender)%>%summarise(
                                 car.md=mean(trc.d),
                                 oth.md=mean(tro.d),
                                 rat.md=mean(tro.d)/mean(trc.d))
```



## 2. Weighting time diary data

### Distribution of days with and without diary weights...


This is a bit more complicated in the  case of MTUS. Let us have a look at the distribution of days by country. 

```{r initial}
round(100*
        prop.table(
          xtabs(~country+day,d)
          ,1),1)
```
Let us now apply some weights:

```{r weight}
round(100*prop.table(xtabs(propwt~country+day,d),1),1)
```
After applying the diary weights, the distribution of days has become roughly equal PROPWT is the diary weight

Note that due to the heterogeneity in the survey design of the studies in the MTUS dataset, the weights do not compensate for month of the year.

```{r month}
round(100*prop.table(xtabs(propwt~country+month,d),1),1)
```


### Another illustration of diary weights from the  the UKTUS 2015

We first load the dataset (this time a time-slot dataset in wide format, with therefore one line per day). This dataset is available to UK-based users registered with the UK Data Service. It could not be shared as part of this training course

```{r uktus1}
sl<-read_dta("data/uktus15_diary_wide.dta")%>% 
    filter(DVAge>=16 & DVAge<65 & IndOut==11)%>%
    select(serial,pnum,daynum,DiaryDay_Act,dmonth,act1_1:act1_144,
          wher_1:wher_144,wit0_1:wit0_144,dia_wt_a,dia_wt_b)
```
`DIA_WT_A` is the diary weight.
`DiaryDay_Act` is the day the diary was recorded (not necessarily the same as the randomly allocated day ).

```{r uktus2}
round(prop.table(xtabs(~DiaryDay_Act,sl))*100,1)
round(prop.table(xtabs(dia_wt_a~DiaryDay_Act,sl))*100,1)
```
By contrast with the MTUS, the UK TUS 2014-15 also compensates fro month of the year ....
```{r uktu3}
round(prop.table(xtabs(~dmonth,sl))*100,1)

round(prop.table(xtabs(dia_wt_a~dmonth,sl))*100,1)
```

### Visualising  the number of episodes in long format episode datasets

This is what a full diary looks like

data.frame(ep)[1:20,]

### And now we compute the number of episodes per person per day
ep<-ep%>%group_by(study,hldid,persid,id)%>%mutate(maxep=max(epnum))%>%ungroup()

### Most diaries have between 15 and 25 episodes 
table(ep$maxep[ep$epnum==1])
hist(ep$maxep[ep$epnum==1],freq = F)

### Diaries with less than 7 episodes
prop.table(
  table(ep$maxep[ep$epnum==1]<7)
)
### More than 80% of diaries have 15 episodes recorded or more
prop.table(
  table(ep$maxep[ep$epnum==1]>=15)
)

### UK diaries have the largest number of episodes recorded on average
ep%>%group_by(study)%>%filter(epnum==1)%>%summarise(epnr=mean(maxep))



## 2. Tempograms

One way of describing how  people spend their time is to compute means and other summary  statistics. Another one is to compare cross-sections of people's activities throughout the day. A relatively straightforward way of doing so is to draw tempograms, which are plots of the proportion (or percentages of people engaged in a series of activities over a 24h period).

In order to plot tempograms, we need to compute for every time point in the diary, the number of people engaged in the activities we are interested in. We can do it for all activities in which people are engaged in, by simplifying and regrouping the detailed codes from their time... Or for a limited range that we select.

For this example, we will be working with the UK Time Use Study from 2014-15. Also, we need to work with time slots data either in long or wide format.
The easiest way of plotting tempograms  is to start from wide data.

We will use a wide time slot dataset derived from the 2014-15 UK Time Use Survey. Variables include: 
  
  - sex (1= Male, 2= Female)
  - wkd weekend vs weekday, as previously
  - act1_1...act1_144 main activity (HETUS Nomenclature, 3 digit) 


```{r opensl}
sl<-data.frame(read_dta("data/tmp_slot.dta"))
```

Let's have a quick look at the data:

```{r looksl}
 dim(sl)
 ls(sl)
 names(sl)
 head(sl)
```
 
This vector will store the plain English version of the categories of activity  we are going to use for the tempograms

```{r labl}
lab.l<-c("Sleep & Self-care","Work & education","Reproductive work",
         "Caring","Volunteering","Physical activity","Leisure","Other")

```
We are going to plot tempograms by gender and weekend vs weekdays.

We first need to create a dataset for  storing the results. We could also have only one such dataset, and add a gender vs weekend/weekday variable

```{r resdat}
wem<-data.frame(slc=0,wrk=0,hwk=0,car=0,vol=0,phy=0,lei=0,oth=0)
wkm<-data.frame(slc=0,wrk=0,hwk=0,car=0,vol=0,phy=0,lei=0,oth=0)
wef<-data.frame(slc=0,wrk=0,hwk=0,car=0,vol=0,phy=0,lei=0,oth=0)
wkf<-data.frame(slc=0,wrk=0,hwk=0,car=0,vol=0,phy=0,lei=0,oth=0)
```

We are now going to loop  through the 144 variables recording primary activities for each time slot of the day 

```{r loopwide}
for(i in paste0("act1_",1:144)){
### For convenience we create a temporary 
### variable which will take the value of the current primary activity 
### for each value of i in the loop
  
  sl$tmp<-sl[,which(names(sl)==i)]
  
  ### We then sequentially recode the primary activities
  ### into our eight category typology
  
  sl$act.f=ifelse(sl$tmp<1000,lab.l[1],"Other")
  sl$act.f=ifelse((sl$tmp>=1000 & sl$tmp<3000) | 
                   (sl$tmp>= 9100 & sl$tmp<9300),lab.l[2],sl$act.f)
  sl$act.f=ifelse((sl$tmp>=3000 & sl$tmp<3800 ) | 
                   sl$tmp ==9360 & sl$tmp <9390,lab.l[3],sl$act.f)
  sl$act.f=ifelse((sl$tmp>=3800 & sl$tmp<4000) | 
                   sl$tmp ==9380,lab.l[4],sl$act.f)
  sl$act.f=ifelse((sl$tmp>=4000 & sl$tmp<5000)  | 
                   (sl$tmp>= 9400 & sl$tmp<9500),lab.l[5],sl$act.f)
  sl$act.f=ifelse(sl$tmp>=6000  & sl$tmp<7000,lab.l[6],sl$act.f)
  sl$act.f=ifelse((sl$tmp>=5000 & sl$tmp<6000) | 
                   (sl$tmp>=7000 & sl$tmp<9000)  |
                    (sl$tmp>= 9400 & sl$tmp==9900),lab.l[7],sl$act.f)
  
  ### We convert  it into a factor, so that we can order 
  ### the categories the way we like
  
  sl$act.f<-as.factor(sl$act.f)
  sl$act.f<-factor(sl$act.f,levels=levels(sl$act.f)[c(6,8,5,1,7,4,2,3)])
  
  ### Finally we compute the proportion of respondents 
  ### in each category at each time slot
  ### for each one the gender by weekend/weekday subpopulations 
  wkm<-rbind(wkm,prop.table(table(sl$act.f[sl$wkd=="Weekday" & sl$sex==1]))*100)
  wem<-rbind(wem,prop.table(table(sl$act.f[sl$wkd=="Weekend" & sl$sex==1]))*100)
  wkf<-rbind(wkf,prop.table(table(sl$act.f[sl$wkd=="Weekday" & sl$sex==2]))*100)
  wef<-rbind(wef,prop.table(table(sl$act.f[sl$wkd=="Weekend" & sl$sex==2]))*100)
  
}    

```
  
  
Note: the same result can be obtained with `xtabs()` using weighted estimates:

```
  wkm<-rbind(wkm,prop.table(xtabs(~act.f,sl%>%filter(wkd=="Weekday" & sex==1))*100)
  wem<-rbind(wem,prop.table(xtabs(~act.f[sl$wkd=="Weekend" & sl$sex==1]))*100)
  wkf<-rbind(wkf,prop.table(xtabs(~act.f[sl$wkd=="Weekday" & sl$sex==2]))*100)
  wef<-rbind(wef,prop.table(xtabs(~act.f[sl$wkd=="Weekend" & sl$sex==2]))*100)
```


A little bit of cleaning up is  necessary. We also need to create the time variable, flagging the time slot

```{r cleantime}

wkm<-wkm[2:nrow(wkm),]
wkm$time<-1:144
wem<-wem[2:nrow(wem),]
wem$time<-1:144

wkf<-wkf[2:nrow(wkf),]
wkf$time<-1:144
wef<-wef[2:nrow(wef),]
wef$time<-1:144
```

We can now create the four plots using a loop as well:

```{r loopplot}
for(z in c("wkm","wem","wkf","wef")) {

### we need to 'melt'ie convert the dataset into a long format version
### in order for it to be processed easily by ggplot
  
tmp.m<-melt(eval(parse(text=z)),id.vars="time") 

### Formatting the time values
### accounting for a start of the day at 4 AM
names(tmp.m)[2]<-"Activity"
tmp.m$min=230+(tmp.m$time*10)
tmp.m$min<-ifelse(tmp.m$min>1440,tmp.m$min-1440,tmp.m$min)
###Clock 
tmp.m$ct=paste0(floor(tmp.m$min/60),":",round(((tmp.m$min/60)-floor(tmp.m$min/60))*60))
levels(tmp.m$Activity)<-lab.l

### Plotting the results

#graphics.off()
tmp.plot<-ggplot(tmp.m, aes(x=time,y=value))+
  geom_area(aes(color=Activity,fill=Activity),alpha=0.5,position='stack')+
  scale_x_continuous(name="Time of the day",breaks=c(seq(1,144,6)),labels=c(4:24,1:3))+
  scale_y_continuous(name="% of  people by activity")+
  #### Not strictly necessary, but makes our plot looks nice
  theme(
        legend.title= element_blank(),
        legend.text= element_text(size=9),
        legend.position="bottom",
        axis.text.y= element_text(size=11),
        axis.text.x= element_text(size=11),
        panel.background = element_rect(fill = "transparent"), 
)

print(tmp.plot)

ggsave(tmp.plot, filename=paste0("temp_",z,".png"),
                 height=8,
                width = 12,bg = "transparent")
}



#summary(glm(main7~age+sex*as.factor(study),d%>%filter(main7>0),family=gaussian()))
```



